---
title: Exploration of the Modern Day NFL
author: ~
date: '2020-12-07'
slug: exploration-of-the-modern-nfl
categories: []
tags: []
---

### Introduction

The NFL Elo dataset contains 256 observations of 30 variables. The data was collected by downloading nfl_elo_latest.csv from the FiveThirtyEight Github page. Altogether, the data comprised in this file details the game-by-game Elo ratings and forecasts for the latest NFL regular season up until Jan 3, 2021. There are many factors that go into these predictive metrics, however the more important ones refer to the QB play on a given team. A few questions that I'm going to answer with the data are as follows: is there a significant gap in a team's win probability once the QB rating is mixed in? Which teams have the biggest differential in this split? Additionally, has there been a steady increase of points scored throughout the regular season? If so, what might be the reasoning behind that? All these questions should help us to understand the true value of the QB in today's NFL as well as where the game is headed as far as scoring is concerned.

### Exploratory Data Analysis 

```{r}
library(tidyverse)
elo_variables_1 <- c("team1", "elo_prob1", "qbelo_prob1")
win_prob_team_1 <- dat_nfl[elo_variables_1] %>% na.omit()
home_team_plot <- ggplot(data = win_prob_team_1, aes(x = elo_prob1, y = qbelo_prob1, color = team1)) + geom_point() + facet_wrap(~ team1) + ggtitle("Home Team Win Probability Based on Elo with and without the QB") + theme(legend.position = "none")
team1_plot
```

The data visualizations above seek to establish a correlation between a team's win probability based on elo ratings and a team's win probability based on team elo ratings adjusted for the QB. Across the board, a positive relationship exists among these two variables. Furthermore, this indicates that if a team has an above-average to good starting QB, their chances of winning minus the signal-caller generally resemble a similar percentage. Next, we will perform some calculations on summary statistics from this data to determine which teams in the NFL have the greatest disparity in win probability when the QB is added to the equation.

```{r}
home_elo_margin <- win_prob_team_1 %>% mutate(elo_diff_1 = elo_prob1 - qbelo_prob1)
home_elo_margin
```